Uncovering underlying musical content in the temporal domain \\ \small Leveraging deep learning, inductive bias, and aural music skills to learn embedding spaces from raw waveforms with applications to MIR downstream tasks

### 2023-05-03 evaluation/subsections/beyond dsp.tex
Updates from Overleaf

\section{Motivation}

Personal, academic, and industrial motivations drive this thesis.

This research is conducted in collaboration with \textit{Epidemic Sound AB (ES)} \cite{EpidemicSite}, a prominent Swedish company that provides a global library of over 40,000+ royalty-free soundtrack music and 90,000+ sound effects, and the \textit{Music Technology Group (MTG)} at Universitat Pompeu Fabra in Barcelona, Spain.

The MTG specializes in developing new technologies related to sound and music, including music information retrieval, digital signal processing, computational models of music perception and cognition, and interactive music systems.

The collaboration with Epidemic Sound and the MTG aims to deepen our understanding of music and its fundamental structures, potentially leading to advancements and benefits for the company's technical products, services, and MIR academic research.

\subsubsection{Personal motivation}

As a music enthusiast, professional musician, and scholar, I am passionately committed to exploring the foundational elements of musical composition, investigating the diverse techniques employed in creating musical works, and deciphering the intricate relationships among them. This passion fuels my quest for knowledge on extracting valuable information embedded within music in any domain, being my main interest in the audio and symbolic domain, contributing to developing more sophisticated and effective music-related technologies and solutions in the industrial landscape.

I advocate for exposure as the primary method of learning music, emphasizing the importance of immersing oneself in various musical experiences to foster a holistic understanding of music as an art form. Individuals can cultivate well-rounded musical expertise by engaging with different styles, genres, techniques, and learning approaches, ultimately promoting creativity and experimentation.

\subsubsection{Academic motivation}

The emergence of self-supervised models that learn embedding spaces for retrieving musical content from audio signals has unveiled new opportunities in academic research. By capturing high-level semantic information from raw audio data in an unsupervised manner, these models can potentially revolutionize various aspects of the music industry. Their applications encompass music information retrieval, sharing learned latent representations for MIR \cite{HamelTransferSimilarity}, and fostering interdisciplinary research and innovation in musicology, computer science, and artificial intelligence.

\subsubsection{Industrial and corporate motivation}

In the industrial and corporate realm, the embedding spaces generated by these models can be utilized to create personalized and intelligent music recommendation systems \cite{Chen2020LearningRecommendation}\cite{epidemic}, facilitate effective enforcement of intellectual property rights, and contribute to the development of more sophisticated audio editing-production tools and technical products \cite{WonEmotionStories}. These advancements ultimately enhance our interaction with music in the digital age, promoting innovation and collaboration across the music industry.

While it is true that the theoretical foundations of machine learning can be traced back to the pioneering work of figures such as Alan Turing and Claude Shannon in the mid-twentieth century, we are currently experiencing a period of significant breakthroughs in AI research \cite{Vaswani2017AttentionNeed}. These breakthroughs have facilitated the implementation of AI in widely used products \cite{OpenAI2023GPT-4Report}. Many researchers and corporations are keen to stay abreast of these developments and contribute to this exciting and rapidly evolving field.

\subsection{About music aural skills and high-level perceptual concepts}

I have consistently been inspired by musicians and researchers who strive to uncover musical ground truth within an existing tonal paradigm, such as Heinrich Schenker and his school of thought, or by challenging it, exemplified by Arnold Schoenberg or George Russell \cite{LydianRussell}. These approaches aim to identify abstract concepts that reinforce or disrupt the tonal foundation. Their ultimate goal is to advance the tonal landscape, providing musicians with a dependable playground for growth, development, and further understanding of tonal paradigms.

%%%%%%%%%%%%%%%%%%

\begin{figure}[ht]
\includegraphics[clip,width=\columnwidth]{figures/schenkerian analysis/SchubertOp4no3.png}% 
\caption{\small{Small excerpt of \textit{Wandrers Nachtlied, Op. 4, D. 224} by Franz Schubert. We can see the passage's original score, the schenkerian unfolding of the melody, the chord degrees analysis, and the tonal function.}}
\label{fig:Wandrers Nachtlied, Op. 4, D. 224}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Schenkerian Analysis, developed by Austrian music theorist Heinrich Schenker, investigates the underlying structure of tonal music by focusing on the hierarchical relationships between pitches. This analytical approach is founded on the premise that all tonal music shares a fundamental structure known as "Ursatz" or "basic shape." The method identifies a piece's stepwise descending line (Urlinie) and bass arpeggiation (Bassbrechung), simplifying it to its essential elements and examining how different layers contribute to the overall structure. Central concepts in Schenkerian Analysis include prolongation levels, harmony, counterpoint, voice-leading, and Ursatz (fundamental structure). Despite criticism for reducing all tonal works to a limited number of background structures, Schenker highlights the significance of individual elaboration in determining a piece's identity and meaning.

While such analysis is designed for the symbolic domain, we argue that musical abstraction is also feasible in the audio time domain. In this section, we explore the application of Schenkerian Analysis principles to the Music Information Retrieval (MIR) domain by examining the foreground, middle-ground, and background levels of musical structure as follows:

\begin{itemize}
\item \textbf{Foreground}: The surface level of a musical composition, consisting of the notes, rhythms, and articulations directly perceived by the listener, representing the most detailed level of the musical structure.

\item \textbf{Middle-ground and mesostructures}: An intermediate level of analysis focusing on essential structural elements such as harmonic progressions, thematic materials, and voice-leading reductions that contribute to the overall composition. In the MIR domain, this level can also be understood as mesostructure. Mesostructures are intermediate levels of articulation between waveshapes' microstructure and musical forms' macrostructure. Examples of mesostructures include melody, arpeggios, syncopation, polyphonic grouping, and textural contrast. Mesostructures might have received limited attention in deep learning, as current models only focus on microstructures \cite{Mesostructures2023}.

\item \textbf{Background}: The most abstract level of musical structure, representing the fundamental core consisting of the \textit{Urlinie} and \textit{Bassbrechung}, which serves as the foundation for the entire composition.
\end{itemize}

By understanding and applying Schenkerian Analysis principles to the MIR domain, we can better understand the complex relationships between different levels of musical structure in the audio time domain, allowing for a more comprehensive analysis of tonal music.

\subsection{Beyond Digital Signal Processing (DSP)}

@@

\subsubsection{Sample MIR task}

>>>>

\section{Beyond Digital Signal Processing (DSP)}

@@

\subsection{Sample MIR task}

### 2023-05-03 introduction/Subsections/Introduction.tex
Updates from Overleaf

Music, an essential component of human culture, history, and society, has existed since ancient times. It reflects the changes in our cultures, traditions, and beliefs as it evolves. Studying music, however, has always been a challenging and complex task due to its subjective nature and the sheer variety of musical styles, genres, and cultures that exist worldwide.

@@

Music similarity is a critical factor in query-by-example retrieval systems. However, similarity calculation and definition can differ depending on user needs and retrieval scenarios. While acoustic properties like voice, genre, melody, tempo, instrumentation, and rhythm contribute to similarity, personal preferences, and cultural context also play a significant role. For instance, a user may avoid a particular artist's music despite its similarity to their favorite artists due to the album cover. Moreover, the user's context, such as driving or running, can also influence their preference and perception of similarity.

Metadata is the traditional approach to electronic music searches, but it can be impractical and unintuitive for users who are unfamiliar with most tracks. Retrieval systems often overlook the cultural, personal, and musical aspects of music information needs, rendering them less useful for non-expert users. Many people search for music based on emotions or usage context rather than bibliographic information.

>>>>

Music \footnote{In this thesis, "music" refers to the Western tradition, characterized by conventions, practices, and aesthetics that have developed primarily in Europe and North America. This usage is not intended to deny the diversity and richness of other musical traditions worldwide but instead reflects the present study's focus on a specific cultural context.}, an essential component of human culture, history, and society, has existed since ancient times. It reflects the changes in our cultures, traditions, and beliefs as it evolves. Studying music, however, has always been a challenging and complex task due to its subjective nature and the sheer variety of musical styles, genres, and cultures that exist worldwide.

@@

Music similarity is a critical factor in query-by-example retrieval systems. However, similarity calculation and definition can differ depending on user needs and retrieval scenarios. While acoustic properties like voice, genre, melody, tempo, instrumentation, and rhythm contribute to similarity, personal preferences, and cultural context also play a significant role. For instance, a user may avoid a particular artist's music despite its similarity to their favorite artists due to the album cover. Moreover, the user's context, such as driving or running, can influence their preference and perception of similarity.

Metadata is the traditional approach to electronic music searches, but it can be impractical and unintuitive for users unfamiliar with most tracks. Retrieval systems often overlook the cultural, personal, and musical aspects of music information needs, rendering them less useful for non-expert users. Many searches for music are based on emotions or usage context rather than bibliographic information.

### 2023-05-03 methods/subsections/machine learning.tex
Updates from Overleaf

\subsection{About ANNs}
The concept of neural networks can be traced back to the early 20th century, with the pioneering work of Warren McCulloch and Walter Pitts in 1943. They proposed a mathematical model of a neuron known as the McCulloch-Pitts neuron, which provided a foundation for future research in artificial intelligence. However, Frank Rosenblatt's invention of the perceptron in 1957 marked the beginning of modern neural networks. The perceptron is a single-layer neural network capable of learning linearly separable patterns.

@@

In a more technical description, an ANN consists of an input layer, one or more hidden layers, and an output layer. Each neuron in a layer receives input from the previous layer and computes an output value using an activation function. Some standard activation functions include the sigmoid, hyperbolic tangent (tanh), and rectified linear unit (ReLU). The output of the last layer represents the prediction or classification result.

@@

Deep learning is a machine learning subfield that utilizes deep neural networks to learn hierarchical data representations automatically. These networks handle complex, high-dimensional data across domains such as computer vision and natural language processing. Their effectiveness stems from their nonlinear modeling capabilities and scalability with large datasets.

>>>>

\subsection{About ANNs}
The concept of neural networks can be traced back to the mid-20th century, with the pioneering work of Warren McCulloch and Walter Pitts in 1943. They proposed a mathematical model of a neuron known as the McCulloch-Pitts neuron, which provided a foundation for future research in artificial intelligence. However, Frank Rosenblatt's invention of the perceptron in 1957 marked the beginning of modern neural networks. The perceptron is a single-layer neural network capable of learning linearly separable patterns.

@@

In a more technical description, an ANN consists of an input layer, one or more hidden layers, and an output layer. Each neuron in a layer receives input from the previous layer and computes an output value using an activation function. The output of the last layer represents the prediction or classification result.

@@

Deep learning is a machine learning subfield that utilizes deep neural networks to learn hierarchical data representations. These networks handle complex, high-dimensional data across several domains. Their effectiveness stems from their nonlinear modeling capabilities and scalability with large datasets.

### 2023-05-03 methods/subsections/model architecture and training strategy.tex
Updates from Overleaf

\section{Model architecture}

@@

Where:

$y_{c_o}(n)$ is the output for the $c_o$-th output channel at position $n$
$C_{\text{in}}$ is the number of input channels
$x_{c_i}(n)$ is the input for the $c_i$-th input channel at position $n$
$w_{c_o, c_i}(k)$ is the kernel weight for the $c_o$-th output channel and $c_i$-th input channel at position $k$
$K$ is the kernel size
The stride is set to 1 and padding is set to 1 (equal padding $\lfloor K/2 \rfloor$ on both sides)
s
    \item Batch normalization. We process a batch of data containing anchor, positive, and negative samples. It pads the waveforms to the same length, performs online triplet mining to find the hardest negatives, and returns the padded anchors, positives, and hardest negatives.

    Let $A$, $P$, and $N$ represent the anchor, positive, and negative samples. The maximum length in the batch is given by:

@@

We have adapted the original architecture for our experiments to our purpose and needs—the weights initialization and the final average pooling step.

>>>>

@@

where:

$y_{c_o}(n)$ is the output for the $c_o$-th output channel at position $n$
$C_{\text{in}}$ is the number of input channels
$x_{c_i}(n)$ is the input for the $c_i$-th input channel at position $n$
$w_{c_o, c_i}(k)$ is the kernel weight for the $c_o$-th output channel and $c_i$-th input channel at position $k$
$K$ is the kernel size
The stride is set to 1 and padding is set to 1 (equal padding $\lfloor K/2 \rfloor$ on both sides)
s
    \item Batch normalization. We process a batch of data containing anchor, positive, and negative samples. We pad the waveforms to the same  maximum length among batch, performs online triplet mining to find the hardest negatives, and returns the padded anchors, positives, and hardest negatives.

    Let $A$, $P$, and $N$ represent the anchor, positive, and negative samples. The maximum array length in the batch is given by:

@@

We have adapted the original architecture for our experiments to our purpose, which consists of the final average pooling step.

\begin{equation}
y_i = \frac{1}{N} \sum_{j=1}^{N} x_{ij}
\end{equation}

\begin{itemize}
  \item \(y_i\) represents the output tensor after the average pooling operation,
  \item \(x_{ij}\) represents the elements in the 2D input tensor,
  \item \(N\) is the number of elements along the second dimension of the input tensor,
  \item \(i\) iterates over the first dimension of the tensors, and
  \item \(j\) iterates over the second dimension of the input tensor.
\end{itemize}
